Typed Functional Genetic Programming

### 2013-07-17 tex/thesis.tex
[tex]

\noindent
Dedication.

@@

Genetic programming (GP) is an AI technique, which falls into 
broader category of evolutionary algorithms --- 
mataheuristic search algorithms inspired by 
biological evolution by natural selection.
I dare say many people perceive its beauty in the fact that GP is
a computer program constructing other computer programs 
%(property sometimes called \textit{automatic programming}) 
with desired properties by breeding them. 
It is also pretty successful technique in a number of areas. 
And perhaps that is why GP has become very popular.

GP constructs the population of solution programs from 
two sets of building blocks; set F of functions 
and set T of constants and variables. 
In classical GP \textit{type consistency} of those 
building blocks is required, property also called \textit{closure}.
This means that all functions must accept

@@

@@

@@

@@

  Henk Barendregt, Wil Dekkers, Richard Statman,
  \emph{Lambda Calculus With Types}.
  Cambridge University Press,
  2010. \\
  \url{http://www.cs.ru.nl/~henk/book.pdf}

>>>>

\noindent
$<$dedication$>$

@@

Genetic programming (GP) is an AI technique, which falls into broader category 
of evolutionary algorithms  ---  mataheuristic  search algorithms inspired  by 
biological evolution by natural selection. I dare say many people perceive its beauty in the  fact that GP is a computer program  constructing other computer programs %(property sometimes called \textit{automatic programming}) 
with desired properties by breeding them. 
It is also pretty successful technique in a number of areas \cite{koza05}. 
And perhaps that is why GP has become very popular.

GP constructs the population of solution programs from 
two sets of building blocks; set $F$ of functions 
and set $T$ of constants and variables. 
In classical GP \textit{closure} of those 
building blocks is required.
This means that all functions in $F$ must accept all 
constants and variables from $T$ as its arguments and produce 
such values that are also acceptable by all functions in $F$ as arguments.
Main motivation behind this thesis is to design and implement a system that
eliminates this constraint. 
This can be achieved by utilization of \textit{types}.\\

Utilization of types in GP is not a new idea; 
it was dealt with in various previous works 
\cite{yu01,montana95,haynes96,olsson94}.
The contribution of this thesis can be seen 
in that it deals with this task using concepts from
theorie of typed $\lambda$-calculus and combinatory logic
and shows how they can be used for benefit in GP.\\

Notable highlights are: 
\begin{itemize}
 \item \textit{Inhabitation tree} (a slight modification 
   of \textit{inhabitation machine} presented in \cite{barendregt10}
   suited better for implementation purposes) 
   is presented and used for \textit{generating of individuals}.
 \item \textit{Abstraction elimination} is utilized
   to enable use of simple \textit{mutation operator}.
 \item \textit{Polymorphic} generalization of inhabitation tree,
   enabling use of polymorphic functions as building blocks, 
   is presented.

\end{itemize}

Design of a GP system utilizing these methods is presented.
This design aims to be generalization of classical GP 
developed by Koza \cite{koza92}, rather than being 
specialized variant of GP using more recent techniques.
This decision is based on belief that modern specialized 
techniques can be incorporated later and on belief
that simplicity of design of classical GP plays nicely
with testing those $\lambda$-calculus concepts in 
the evolutionary context.

The implementation aims

@@

& zmínit že existuje i jinej pohled na to proč tam zavádět typy:
  && mužem koukat na to tak že konstrejnt je naše TuF tim, že 
     v ní nemužou bejt libovolný funkce
  && ale tuhle constrain mužeme omezit tim že konstrjnujeme 
     právě použití tech bilding bloku, kde constrains jsou daný těma typama
     - jakoby výměna global constrain za localní konstrejny na tech bilding blockách
  && druhá možná ješte silnější motivace je že pomocí typu pak mohou vznikat 
     smysluplnější programy, tím že dostavají vstupy které přesně čekají
     a dávají výstupy které mají nějaké očekávané vlastnosti

@@

% TODO bacha neni tam jednotnej formát!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

@@

\bibitem{koza05}

  Koza, J.R., Keane, M., Streeter, M., Mydlowec, W.,Yu, J., Lanza, G. 
  \emph{Genetic Programming IV: Routine Human-Competitive Machine Intelligence.} 
  Springer, 2005. ISBN 978-0-387-26417-2

\bibitem{yu01}

  T. Yu. 
  \emph{Hierachical processing for evolving recursive and modular 
        programs using higher order functions and lambda abstractions}. 
  Genetic Programming and Evolvable Machines,
  2(4):345–380, December 2001. ISSN 1389-2576.

\bibitem{montana95}

D. J. Montana. 
\emph{Strongly typed genetic programming.} 
Evolutionary Computation, 3(2): 199–230, 1995.
%URL \url{ http://vishnu.bbn.com/papers/stgp.pdf }. nefacha

\bibitem{haynes96}

T. D. Haynes, D. A. Schoenefeld, and R. L. Wainwright. 
\emph{Type inheritance in strongly typed genetic programming.} 
In P. J. Angeline and K. E. Kinnear, Jr., editors, Advances
in Genetic Programming 2, chapter 18, pages 359–376.
MIT Press, Cambridge, MA, USA, 1996. ISBN 0-262-01158-1.\\ 
URL 
\url{http://www.mcs.utulsa.edu/~rogerw/papers/Haynes-hier.pdf}.

\bibitem{olsson94}

J. R. Olsson. 
\emph{Inductive functional programming using incremental program 
transformation and Execution of logic programs by 
iterative-deepening A* SLD-tree search.} 
Dr scient thesis, University of Oslo, Norway, 1994.

@@

  Henk Barendregt, Wil Dekkers, Richard Statman,
  \emph{Lambda Calculus With Types}.
  Cambridge University Press,
  2010. \\
  URL \url{http://www.cs.ru.nl/~henk/book.pdf}.
